---
title: "clean-GCCR005"
author: "Pari"
date: '2022-03-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import dataset
```{r import, include = FALSE}
library(readr)
data <- read_csv("GCCR_dataset_parent.csv")
View(data)
str(data)
```

# Count participants who have ever accessed a session.
118 unique participants. 
108 accessed 1 session.
8 accessed 2 sessions.
1 accessed 5 sessions.
1 accessed 9 sessions.
```{r select}
install.packages("dplyr")
library(dplyr)
count_table<- table(data %>% group_by(subjectid) %>% summarize(count=n()))
data2<-as.data.frame(count_table)
table(data2$Freq, data2$count)
```

# Select only participants who provided consent, and count how many accessed sessions
48 unique participants, who accessed 1 session.
39 unique participants qualify. 
```{r select_consent}
data_consent<-subset(data, consent == 1) 
str(data_consent)
count_table_consent<- table(data_consent %>% group_by(subjectid) %>% summarize(count=n()))
data2_consent<-as.data.frame(count_table_consent)
table(data2_consent$Freq, data2_consent$count)

data_qualify<- subset(data_consent, qualify == 1) 
str(data_qualify)
count_table_qualify<- table(data_qualify %>% group_by(subjectid) %>% summarize(count=n()))
data2_qualify<-as.data.frame(count_table_qualify)
table(data2_qualify$Freq, data2_qualify$count)

```
# Describe the sample which consented and qualified
39 unique participants with age data. Mean: 40.05, sd: 12.32; 
32 women, 6 men, 1 other

By looking at the region column, we can see that the vast majority of participants are from the US.

13 unique participants are white, 1 prefers not to answer and 4 are Other. 

19 never smokers, 5 currently smoke, 11 smoked in the past, 2 prefer not to say. 

14 none
6 high blood pressure
1 heart disease
0  diabetes
3 obesity
0 lung disease
0 head trauma
1 neurological disease
0 cancer + chemo
0 cancer - chemo
1 chronic sinus problems
0 pre-COVID smell disorders
0 pre-COVID taste disorders
11 seasonal allergies
0 Other
```{r describe_consent}
#age
data_qualify$age<-2021-data_qualify$born
str(data_qualify)
mean(data_qualify$age, na.rm = TRUE)
sd(data_qualify$age, na.rm = TRUE)

#gender
table(data_qualify$gender)

#ethnicity
table(data_qualify$ethnicity)

#smoker
table(data_qualify$smoker_v2)

#pre-existing conditions
table(data_qualify$conditions_v2___0)
table(data_qualify$conditions_v2___1)
table(data_qualify$conditions_v2___2)
table(data_qualify$conditions_v2___3)
table(data_qualify$conditions_v2___4)
table(data_qualify$conditions_v2___5)
table(data_qualify$conditions_v2___6)
table(data_qualify$conditions_v2___7)
table(data_qualify$conditions_v2___8)
table(data_qualify$conditions_v2___9)
table(data_qualify$conditions_v2___10)
table(data_qualify$conditions_v2___11)
table(data_qualify$conditions_v2___12)
table(data_qualify$conditions_v2___13)
table(data_qualify$conditions_v2___88)
```
# describe COVID diagnosis

One person did not respond. 36 unique participants had COVID-19 and 2 had not; 31 had a positive lab test, 1 were diagnosed by medics; 1 was not diagnosed by a medical professional; 2 were unknown.
```{r covid}
table(data_qualify$covid_diagnosis_v2)
table(data_qualify$covid_diagnosis_how_v2)

data_qualify_covid<-subset(data_qualify, covid_diagnosis_v2 == 1) 
str(data_qualify_covid)
table(data_qualify_covid$covid_diagnosis_how_v2)
```

# Symptoms during COVID

0 No Symptoms
20 Fever
16 Dry Cough
9 Cough with mucus 
9 Shortness of breath
12 Chest tightness
19 Runny nose
16 Sore throat
23 Changes in food flavor
34 Changes in smell
30 Changes in taste
17 Loss of appetite
27 Headache
27 Body aches
28 Fatigue
12 Diarrhea
5 Abdominal pain
8 Nausea
6 Skin sensitivity
13 Dry mouth
8 Other
```{r covid_symptoms}
table(data_qualify$resp_sypmtoms_v2___0)
table(data_qualify$resp_sypmtoms_v2___1)
table(data_qualify$resp_sypmtoms_v2___2)
table(data_qualify$resp_sypmtoms_v2___3)
table(data_qualify$resp_sypmtoms_v2___4)
table(data_qualify$resp_sypmtoms_v2___5)
table(data_qualify$resp_sypmtoms_v2___6)
table(data_qualify$resp_sypmtoms_v2___7)
table(data_qualify$resp_sypmtoms_v2___8)
table(data_qualify$resp_sypmtoms_v2___9)
table(data_qualify$resp_sypmtoms_v2___10)
table(data_qualify$resp_sypmtoms_v2___11)
table(data_qualify$resp_sypmtoms_v2___12)
table(data_qualify$resp_sypmtoms_v2___13)
table(data_qualify$resp_sypmtoms_v2___14)
table(data_qualify$resp_sypmtoms_v2___15)
table(data_qualify$resp_sypmtoms_v2___16)
table(data_qualify$resp_sypmtoms_v2___17)
table(data_qualify$resp_sypmtoms_v2___18)
table(data_qualify$resp_sypmtoms_v2___19)
table(data_qualify$resp_sypmtoms_v2___88)
```

# Baseline
```{r baseline}

str(data_qualify)
mean(data_qualify$smell_today, na.rm = T) 
mean(data_qualify$smell_6mo, na.rm = T)
sd(data_qualify$smell_today, na.rm = T)
sd(data_qualify$smell_6mo, na.rm = T)


mean(data_qualify$blocked_today, na.rm = T)
mean(data_qualify$blocked_6mo, na.rm = T)
sd(data_qualify$blocked_today, na.rm = T)
sd(data_qualify$blocked_6mo, na.rm = T)

mean(data_qualify$taste_today, na.rm = T)
mean(data_qualify$taste_6mo, na.rm = T)
sd(data_qualify$taste_today, na.rm = T)
sd(data_qualify$taste_6mo, na.rm = T)

mean(data_qualify$oralirr_today, na.rm = T)
mean(data_qualify$oralirr_6mo, na.rm = T)
sd(data_qualify$oralirr_today, na.rm = T)
sd(data_qualify$oralirr_6mo, na.rm = T)
#why didnt she made another 4 columns for smell, taste,blocked and oralirr change

install.packages("ggplot2")
library(ggplot2)
# Basic density
p <- ggplot(data_qualify, aes(x=smell_6mo)) + 
  geom_density()
q <- ggplot(data_qualify, aes(x=blocked_6mo)) + 
  geom_density()
r <- ggplot(data_qualify, aes(x=taste_6mo)) + 
  geom_density()
s <- ggplot(data_qualify, aes(x=oralirr_6mo)) + 
  geom_density()

a <- ggplot(data_qualify, aes(x=smell_today)) + 
  geom_density()
b <- ggplot(data_qualify, aes(x=blocked_today)) + 
  geom_density()
c <- ggplot(data_qualify, aes(x=taste_today)) + 
  geom_density()
d <- ggplot(data_qualify, aes(x=oralirr_today)) + 
  geom_density()

```
#smell_change
```{r changed during 6 month}
data_qualify_covid <- data_qualify_covid[data_qualify_covid$subjectid != 101,] # I wanted to delete this row
#add columns for smell, taste, blocked and oralirr change 
data_qualify_covid$smell_change <- data_qualify_covid$smell_today - data_qualify_covid$smell_6mo 
data_qualify_covid$blocked_change <- data_qualify_covid$blocked_today - data_qualify_covid$blocked_6mo 
data_qualify_covid$taste_change <- data_qualify_covid$taste_today - data_qualify_covid$taste_6mo
data_qualify_covid$oralirr_change <- data_qualify_covid$oralirr_today - data_qualify_covid$oralirr_6mo

percentage_change <- function(previous, new, as_decimal = FALSE) {
  x <- ((new - previous) / previous) * 100
  if (as_decimal) x <- x / 100
  return(x)
}
#here they have named the function as percentage_change and defined its parameters inside the curly bracket and used the name of the function

data_qualify_covid$smell_change_p <-
  percentage_change(data_qualify_covid$smell_6mo,data_qualify_covid$smell_today )


data_qualify_covid$blocked_change_p <- percentage_change(data_qualify_covid$blocked_6mo,data_qualify_covid$blocked_today)


data_qualify_covid$taste_change_p <- 
  percentage_change(data_qualify_covid$taste_6mo,data_qualify_covid$taste_today)


data_qualify_covid$oralirr_change_p <- percentage_change(data_qualify_covid$oralirr_6mo,data_qualify_covid$oralirr_today)

mean(data_qualify_covid$smell_change_p, na.rm = T)
sd(data_qualify_covid$smell_change_p, na.rm = T)


e <- ggplot(data_qualify_covid, aes(x=smell_change_p)) + 
  geom_density(fill = 'cyan', color = 'cyan')+
  labs(title = 'change in smell from 6 month ago till now') +
  theme(text = element_text(family = 'Gill Sans', color = "#444444", size = 5)
        ,panel.background = element_rect(fill = '#444B5A')
        ,panel.grid.minor = element_line(color = '#4d5566')
        ,panel.grid.major = element_line(color = '#586174')
        ,plot.title = element_text(size = 8)
        ,axis.title = element_text(size = 8, color = '#DDAA33')
        ,axis.title.y = element_text(vjust = .5, angle = 0)
        ,axis.title.x = element_text(hjust = .5)
  ) 
f <- ggplot(data_qualify_covid, aes(x=blocked_change_p)) + 
  geom_density(fill = 'cyan', color= 'cyan')+labs(title = 'change in blocked nose') +
  theme(text = element_text(family = 'Gill Sans', color = "#444444", size = 5)
        ,panel.background = element_rect(fill = '#444B5A')
        ,panel.grid.minor = element_line(color = '#4d5566')
        ,panel.grid.major = element_line(color = '#586174')
        ,plot.title = element_text(size = 8)
        ,axis.title = element_text(size = 8, color = '#DDAA33')
        ,axis.title.y = element_text(vjust = .5, angle = 0)
        ,axis.title.x = element_text(hjust = .5)
  ) 
g <- ggplot(data_qualify_covid, aes(x=taste_change_p)) + 
  geom_density(fill = 'cyan')+labs(title = 'change in taste from 6 month ago till now') +
  theme(text = element_text(family = 'Gill Sans', color = "#444444", size = 5)
        ,panel.background = element_rect(fill = '#444B5A')
        ,panel.grid.minor = element_line(color = '#4d5566')
        ,panel.grid.major = element_line(color = '#586174')
        ,plot.title = element_text(size = 8)
        ,axis.title = element_text(size = 8, color = '#DDAA33')
        ,axis.title.y = element_text(vjust = .5, angle = 0)
        ,axis.title.x = element_text(hjust = .5)
  ) 
h <- ggplot(data_qualify_covid, aes(x=oralirr_change_p)) + 
  geom_density(fill = 'cyan')+labs(title = 'change in trigeminal from 6 month ago till now') +
  theme(text = element_text(family = 'Gill Sans', color = "#444444", size = 5)
        ,panel.background = element_rect(fill = '#444B5A')
        ,panel.grid.minor = element_line(color = '#4d5566')
        ,panel.grid.major = element_line(color = '#586174')
        ,plot.title = element_text(size = 8)
        ,axis.title = element_text(size = 8, color = '#DDAA33')
        ,axis.title.y = element_text(vjust = .5, angle = 0)
        ,axis.title.x = element_text(hjust = .5)
  ) 

install.packages("ggpubr")
library("ggpubr")
ggarrange(p, q, r, s, 
          ncol = 2, nrow = 2)

ggarrange(a, b, c, d, 
          ncol = 2, nrow = 2)

ggarrange(e, f, g, h, 
          ncol = 2, nrow = 2)
```
#severity
we did median split to convert the continuous data(number of symptoms for each) to categorical 
We have to category of severity: if the number of symptoms are between 1-9 we call them L(less severe) and otherwise is S
severity	/ number of participants: L=Less severe and S= severe
L	20
S	16
```{r severity of covid}
# create a new coulmn for symptoms frequency
data_qualify_covid$symptomfreq <- ""
data_qualify_covid$symptomfreq <- rowSums(data_qualify_covid[ , c(44:63)], na.rm=TRUE)
data_qualify_covid$symptomfreq
#create a new column for severity
data_qualify_covid$severity <- ""

data_qualify_covid$severity[data_qualify_covid$symptomfreq <= 9] <- "L"
data_qualify_covid$severity[data_qualify_covid$symptomfreq > 9] <- "S"                            
data_qualify_covid$severity             
data_qualify_covid

#make a new data set with all the columns that interest us for analysis the data related to smell cange and symptoms
p_dataset <- data_qualify_covid[c('subjectid', 'symptomfreq', 'severity','smell_change_p', 'blocked_change_p', 'taste_change_p', 'oralirr_change_p', 'smell_change', 'blocked_change', 'taste_change', 'oralirr_change')]

# clean the data
#pt 0 for inf NAN...
p_dataset <- p_dataset %>% mutate_if(is.numeric, function(x) ifelse(is.infinite(x), 0, x))
#p_dataset <- replace(p_dataset, is.na(p_dataset),0 )


#mean and sd for changes
library(dplyr)

mean(p_dataset$smell_change_p, na.rm = T)
sd(p_dataset$smell_change_p, na.rm = T)


mean(p_dataset$taste_change_p, na.rm = T)
sd(p_dataset$taste_change_p, na.rm = T)


mean(p_dataset$oralirr_change_p, na.rm = T)
sd(p_dataset$oralirr_change_p, na.rm = T)


mean(p_dataset$blocked_change_p, na.rm = T)
sd(p_dataset$blocked_change_p, na.rm = T)


#mixmodel
#we used a mix model to see the correlation between severity and changes in smell/taste/blocked/oralirr
#there is no significant correlation. The only significant one is the Trigeminal change (oralirr) with the p-value: 0.0412.
lm1 <- lm(smell_change_p ~ severity, p_dataset)
lm2 <- lm(blocked_change_p ~ severity, p_dataset)
lm3 <- lm(taste_change_p ~ severity, p_dataset)
lm4 <- lm(oralirr_change_p ~ severity, p_dataset)

#look at the summary
summary(lm1)
summary(lm2)
summary(lm3)
summary(lm4)
```
#severity pie chart
```{r piechart}
# Pie Chart with Percentages
slices <- c(20,16)

lbls <- c("Less severe","Sever")

pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
pie(slices,labels = lbls, col=rainbow(length(lbls)),main="Symptoms frequency")
```

```{R symptoms freq}
#doughnut piechaet for symptoms freq
slices1 <- c(34, 30, 28, 27, 27, 23, 20, 19, 17, 16, 16, 16, 13, 12, 12, 9, 8, 8, 6, 5)
names1 <- c("Changes in or loss of smell", "Changes in or loss of taste", "Fatigue", "Body ache", "Headache", "Change in food flavour", "Feaver", "Runny nose", "Loss of appetite", "Dry cough", "Couch with mucous", "Sore throat" , "Dry mouth", "Chest tightness", "Diarrhoea", "Difficulty breathing", "Nausea", "Other", "Skin sinsitivity", "Abdominal pain" )
pct1 <- round(slices1/sum(slices1)*100)
lbls <- paste(names1, pct1)
data_doughnut <- data.frame(slices1, names1)

donut <-  ggplot(data = data_doughnut, aes(x=2, y = pct1, fill = color))+
       geom_col(color = "black") +
       coord_polar("y", start = 0) + 
       geom_text(aes(label = paste0(pct1, "%")), 
                          position = position_stack(vjust = 0.5)) +
       theme(panel.background = element_blank(),
             axis.line = element_blank(),
             axis.text = element_blank(),
             axis.ticks = element_blank(),
             axis.title = element_blank(), 
             plot.title = element_text(hjust = 0.5, size = 18)) +
       ggtitle("Symptoms frequency") +
       scale_fill_manual(values = pct1) +
       xlim(0.5, 2.5)


```

#see the correlation between intensity and self-reported olfactory
```{r intensity}


res <- cor.test(data_qualify_covid$oralirritation_intensity, data_qualify_covid$oralirr_today, 
                method = "pearson", use = "complete.obs")
res
res1 <- cor.test(data_qualify_covid$smell_cosmetics_intensity, data_qualify_covid$smell_today, 
                 method = "pearson", use = "complete.obs")
res1
res2 <- cor.test(data_qualify_covid$smell_spice_intensity, data_qualify_covid$smell_today, 
                 method = "pearson", use = "complete.obs")
res2
res <- cor.test(data_qualify_covid$taste_sweet_intensity, data_qualify_covid$taste_today, 
                method = "pearson", use = "complete.obs")
res
res <- cor.test(data_qualify_covid$taste_salt_intensity, data_qualify_covid$taste_today, 
                method = "pearson", use = "complete.obs")
res
res <- cor.test(data_qualify_covid$taste_sour_intensity, data_qualify_covid$taste_today, 
                method = "pearson", use = "complete.obs")
res
res <- cor.test(data_qualify_covid$taste_bitter_intensity, data_qualify_covid$taste_today, 
                method = "pearson", use = "complete.obs")
res



#the oral intensity mean and sd:
mean(data_qualify_covid$oralirritation_intensity, na.rm = T)
sd(data_qualify_covid$oralirritation_intensity, na.rm = T)
#smell intensity
mean(data_qualify_covid$smell_cosmetics_intensity, na.rm = T)
sd(data_qualify_covid$smell_cosmetics_intensity, na.rm = T)

mean(data_qualify_covid$smell_spice_intensity, na.rm = T)
sd(data_qualify_covid$smell_spice_intensity, na.rm = T)

#taste intensity
data_qualify_covid$taste_sweet_intensity
mean(data_qualify_covid$taste_sweet_intensity, na.rm = T)
sd(data_qualify_covid$taste_sweet_intensity, na.rm = T)
data_qualify_covid$taste_salt_intensity
mean(data_qualify_covid$taste_salt_intensity, na.rm = T)
sd(data_qualify_covid$taste_salt_intensity, na.rm = T)
data_qualify_covid$taste_sour_intensity
mean(data_qualify_covid$taste_sour_intensity, na.rm = T)
sd(data_qualify_covid$taste_sour_intensity, na.rm = T)
data_qualify_covid$taste_bitter_intensity
mean(data_qualify_covid$taste_bitter_intensity, na.rm = T)
sd(data_qualify_covid$taste_bitter_intensity, na.rm = T)



# pearson

#correlation between taste intensity and severity
res <- cor.test(data_qualify_covid$oralirritation_intensity, data_qualify_covid$symptomfreq, 
                method = "pearson", use = "complete.obs")
res
res1 <- cor.test(data_qualify_covid$smell_cosmetics_intensity, data_qualify_covid$symptomfreq, 
                 method = "pearson", use = "complete.obs")
res1
res2 <- cor.test(data_qualify_covid$smell_spice_intensity, data_qualify_covid$symptomfreq, 
                 method = "pearson", use = "complete.obs")
res2
res <- cor.test(data_qualify_covid$taste_sweet_intensity, data_qualify_covid$symptomfreq, 
                method = "pearson", use = "complete.obs")
res
res <- cor.test(data_qualify_covid$taste_salt_intensity, data_qualify_covid$symptomfreq, 
                method = "pearson", use = "complete.obs")
res
res <- cor.test(data_qualify_covid$taste_sour_intensity, data_qualify_covid$symptomfreq, 
                method = "pearson", use = "complete.obs")
res
res <- cor.test(data_qualify_covid$taste_bitter_intensity, data_qualify_covid$symptomfreq, 
                method = "pearson", use = "complete.obs")
res


#plot
#bar plot for severity and bitter
barplot <-ggplot(data_qualify_covid, aes(fill=severity, y=taste_bitter_intensity, x=symptomfreq)) + 
  geom_bar(position="dodge", stat="identity")
barplot
#bar plot for severity and bitter
barplot <-ggplot(data_qualify_covid, aes(fill=severity, y=taste_salt_intensity, x=symptomfreq)) + 
  geom_bar(position="dodge", stat="identity")
barplot
#bar plot for severity and bitter
barplot <-ggplot(data_qualify_covid, aes(fill=severity, y=taste_sour_intensity, x=symptomfreq)) + 
  geom_bar(position="dodge", stat="identity")
barplot
#bar plot for severity and bitter
barplot <-ggplot(data_qualify_covid, aes(fill=severity, y=taste_sweet_intensity, x=symptomfreq)) + 
  geom_bar(position="dodge", stat="identity")
barplot
```
# Multiple plots for severity and chemosensory changes
```{r plot}
           

 
p1 <- ggscatter(p_dataset,
          x = "symptomfreq",
          y = "smell_change_p",
          xlab = "Symptoms frequency", 
          ylab = "Smell change",
          color = "severity",
          add = "reg.line") +
  stat_cor(aes(color = severity),method = "spearman", label.x = 3) 

p2 <-  ggscatter(p_dataset,
          x = "symptomfreq",
          y = "taste_change_p",
          xlab = "Symptoms frequency", 
          ylab = "Taste change",  
          color = "severity",
          add = "reg.line") +
  stat_cor(aes(color = severity),method = "spearman", label.x = 3) 

p3 <-  ggscatter(p_dataset,
          x = "symptomfreq",
          y = "blocked_change_p",
          xlab = "Symptoms frequency", 
          ylab = "Blocked change",   
          color = "severity",
          add = "reg.line") +
  stat_cor(aes(color = severity),method = "spearman", label.x = 3) 


p4 <-  ggscatter(p_dataset,
          x = "symptomfreq",
          y = "oralirr_change_p",
          color = "severity",
          xlab = "Symptoms frequency", 
          ylab = "Oral irritation change",   
          add = "reg.line") +
  stat_cor(aes(color = severity),method = "spearman", label.x = 3)

ggarrange(p1, p2,p3,p4, ncol = 2, nrow = 2)  

```